Automated assessment of residual plots with computer vision models

### 2024-02-15 paper/paper.tex
update paper

@@

\articletype{}

@@

\newpage
\tableofcontents
\newpage

@@

However, as pointed out by \citet{li2023plot}, a primary limitation of
the lineup protocol lies in its reliance on human judgments. Unlike
conventional statistical tests that can be performed computationally in
statistical software, the lineup protocol requires human evaluation of
images. This characteristic makes it less suitable for large-scale
applications, given the associated high labor costs and time
requirements.

@@

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
focusing on scatterplot diagnostics. \citet{wilkinson2005graph} expanded
on this work, introducing graph theoretic scagnostics, which encompassed
computable measures applied to planar proximity graphs. These measures,
including, but not limited to, ``Outlying'', ``Skinny'', ``Stringy'',
``Straight'', ``Monotonic'', ``Skewed'', ``Clumpy'', and ``Striated''
aimed to characterize outliers, shape, density, trend, coherence and
other characteristics of the data. While this approach has been
inspiring, there is a recognition \citep{buja2009statistical} that it
may not capture all the necessary visual features that differentiate
actual residual plots from null plots. A more promising alternative
entails enabling machines to learn the function for extracting visual
features from residual plots. Essentially, this means empowering
computers to discern the crucial visual features for residual
diagnostics and determining the method to extract them.

@@

\caption{An example residual vs fitted values plot (red line indicates 0). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity.}\label{fig:false-finding}
\end{figure}

@@

Another design involves using an image as input. The primary advantage
of this design, as opposed to the vector format, is the availability of
the existing and sophisticated image processing architectures developed
over the years, such as the VGG16 architecture proposed in
\citet{simonyan2014very}. These architectures can effectively capture
and summarize spatial information from nearby pixels, which is less
straightforward with vector input. The main considerations are the image
resolution and the aesthetics of the residual plot. In general, higher
resolution provides more information to the model but comes with the
trade-off of increased complexity and greater difficulty in training. As
for the aesthetics of the residual plot, a practical solution is to
consistently present residual plots in the same style to the model. This
implies that the model can not accept arbitrary images as input but
requires the use of the same preprocessing pipeline to convert residuals
and fitted values into a standardized-style residual plot.

@@

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but not included in \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, \(P\) will usually be skewed and
has a long tail.

@@

We tried to train a computer vision model to approximate this distance
measure with a residual plot. Let \(D\) be the result of equation
\ref{eq:kl-0} indicating the extent of the model violations, and our
estimator \(\hat{D}\) is formulated as

@@

With the approximated distance \(\hat{D}\), we will be able to know how
different the underlying distribution of the residuals is from a good
residual distribution. It also provides information for the strength of
the visual signal embedded in the residual plot.

@@

@@

Non-normality violations arise from specifying a non-normal distribution
for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
distinct error distributions are considered, including discrete,
uniform, normal, and lognormal distributions, as presented in Figure
\ref{fig:different-e}. Each distribution imparts unique characteristics
to the residuals. The discrete error distribution introduces
discreteness in residuals, while the lognormal distribution typically
yields outliers. Uniform error distribution may result in residuals
filling the entire space of the residual plot. All of these
distributions exhibit visual distinctions from the normal error
distribution.

@@

For each generated residual plot, we computed four scagnostics --
``Monotonic,'' ``Sparse,'' ``Splines,'' and ``Striped'' -- using the
cassowaryr R package \citep{mason2022cassowaryr}. These computed
measures, along with the number of observations from the fitted model,
were provided as the second input for the computer vision model.
Although other scagnostics are informative, they are currently
unavailable due to a fatal bug in the compiled C program of the interp R
package \citep{Albrecht2023interp}, which may unpredictably crash the
process. For reproducibility, we excluded these scagnostics from the
training data.

@@

The model begins with an input layer of shape
\(n \times h \times w \times 3\), capable of handling \(n\) RGB images.
This is followed by a grayscale conversion layer utilizing the luma
formula under the Rec. 601 standard, which converts the color image to
grayscale. Grayscale suffices for our task since data points are plotted
in black. We experiment with three combinations of \(h\) and \(w\):
\(32 \times 32\), \(64 \times 64\), and \(128 \times 128\), aiming to
achieve sufficiently high image resolution for the problem at hand.

@@

To achieve an optimal deep learning model, hyperparameters like the
learning rate often need to be fine-tuned using a tuner. In our study,
we utilized the Bayesian optimization tuner from the KerasTuner Python
library \citep{omalley2019kerastuner} for this purpose. A comprehensive
list of hyperparameters is provided in Table \ref{tab:hyperparameter}.

@@

\begin{itemize}
\item
  RMSE for the test set
\item
  R\^{}2
\item
  Mean bias deviation to understand the overall bias
\item
  quantile loss to understand how well the model captures the entire
  distribution of D
\item
  Percentage of predictions within a tolerance interval (like 0.1)
\item
  comparison with human visual inference
\item
  overview of the human subject experiment
\item
  metrics
\item
  agreement rate
\item
  check cases that are aggreed
\item
  check cases that are disaggreed
\end{itemize}

@@

The figure highlights that most under-predictions occurred when
\(\hat{D} < 3\), while over-predictions occurred predominantly when
\(3 < \hat{D} < 6\). To comprehend the reasons behind these deviations,
we conducted a meta-analysis on \(D - \hat{D}\) for all factors used in
the data generating process. A regression tree was constructed using the
rpart R package \citep{Terry2022rpart}. The results of the regression
tree are provided in the Appendix.

The meta-analysis revealed that most issues arose from non-linearity
problems and the presence of a second predictor in the regression model.
When the error distribution had a very small variance, all computer
vision models tended to under-predict the distance. Conversely, when the
error distribution had a large variance, all models tended to
over-predict the distance. Therefore, Additionally, for input images
representing null plots, it was expected that the models will
over-predict the distance, as explained in Section
\ref{null-distribution-of-the-approximated-distance}.

@@

\hypertarget{comparison-with-human-visual-inference}{%
\subsection{Comparison with human visual
inference}\label{comparison-with-human-visual-inference}}

@@

For each lineup used in \citet{li2023plot}, there is one actual residual
plot and 19 null plots. While the distance \(D\) for the actual residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to predict \(\hat{D}\) for both the actual residual plots
and the null plots.

@@

Figure \ref{fig:human-mosaic} and Table \ref{tab:human-table} provide
insights into the agreement between decisions made by the computer
vision model and visual tests conducted by humans. It can be observed
that visual tests are less sensitive than the computer vision model, as
they reject almost only when the computer vision model does. The
agreement rates between visual tests and the computer vision model are
67.96\% and 66.84\% for residual plots containing heteroskedasticity and
non-linearity patterns, respectively.

When plotting the decision against the distance, as depicted in Figure
\ref{fig:power}, we observe that for the computer vision model, when the
distance \(D > 3\), almost all residual plots will be rejected. However,
for visual tests conducted by humans, the threshold is higher at
\(D > 5\). Visual tests also demonstrate low sensitivity to residual
plots with small distances compared to the computer vision model.

@@

Figure \ref{fig:delta} indicates that the weighted detection rate
increases as the \(\delta\)-difference increases, particularly when the
\(\delta\)-difference is greater than zero. A negative
\(\delta\)-difference suggests that there is at least one null plot in
the lineup with a stronger visual signal than the actual residual plot.
In some instances, the weighted detection rate is close to one, yet the
\(\delta\)-difference is negative. This discrepancy implies that the
distance measure, or the approximated distance, may not perfectly
reflect actual human behavior.

@@

\caption{\label{tab:human-table}Summary of decisions made by the visual test conducted by human and decisions made by the computer vision model.}
\centering
\begin{tabular}[t]{llll}
\toprule
Violations & \#Samples & \#Agreements & Agreement rate\\
\midrule
heteroskedasticity & 540 & 367 & 0.6796\\
non-linearity & 576 & 385 & 0.6684\\
\bottomrule
\end{tabular}
\end{table}

@@

@@

\caption{Comparison of decisions made by visual tests and the computer vision model. The data points are fitted with logistic regression models with no intercept but an offset equals to $log(0.05/0.95)$.}\label{fig:power}
\end{figure}

@@

\hypertarget{boston-housing}{%
\subsubsection{Boston Housing}\label{boston-housing}}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-8-1} \end{center}

\hypertarget{advertising}{%
\subsubsection{Advertising}\label{advertising}}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-12-1} \end{center}

\hypertarget{wine-quality}{%
\subsubsection{Wine Quality}\label{wine-quality}}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-16-1} \end{center}

@@

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-20-1} \end{center}

>>>>

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

\articletype{Draft paper (Incomplete)}

@@

@@

However, as pointed out by \citet{li2023plot}, a primary limitation of
the lineup protocol lies in its reliance on human judgments. Unlike
conventional statistical tests that can be performed computationally in
statistical software, the lineup protocol requires human evaluation of
images. This characteristic makes it less suitable for large-scale
applications, given the associated high labour costs and time
requirements.

@@

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
focusing on scatter plot diagnostics. \citet{wilkinson2005graph}
expanded on this work, introducing graph theoretic scagnostics, which
encompassed computable measures applied to planar proximity graphs.
These measures, including, but not limited to, ``Outlying'', ``Skinny'',
``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'', ``Clumpy'', and
``Striated'' aimed to characterize outliers, shape, density, trend,
coherence and other characteristics of the data. While this approach has
been inspiring, there is a recognition \citep{buja2009statistical} that
it may not capture all the necessary visual features that differentiate
actual residual plots from null plots. A more promising alternative
entails enabling machines to learn the function for extracting visual
features from residual plots. Essentially, this means empowering
computers to discern the crucial visual features for residual
diagnostics and determining the method to extract them.

@@

\caption{An example residual vs fitted values plot (red line indicates 0). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\% significance level ($p\text{-value} = 0.046$).}\label{fig:false-finding}
\end{figure}

@@

Another design involves using an image as input. The primary advantage
of this design, as opposed to the vector format, is the availability of
the existing and sophisticated image processing architectures developed
over the years, such as the VGG16 architecture proposed in
\citet{simonyan2014very}. These architectures can effectively capture
and summarize spatial information from nearby pixels, which is less
straightforward with vector input. The main considerations are the image
resolution and the aesthetics of the residual plot. In general, higher
resolution provides more information to the model but comes with the
trade-off of increased complexity and greater difficulty in training. As
for the aesthetics of the residual plot, a practical solution is to
consistently present residual plots in the same style to the model. This
implies that the model can not accept arbitrary images as input but
requires the use of the same pre-processing pipeline to convert
residuals and fitted values into a standardized-style residual plot.

@@

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but not included in \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate log-normal distribution, \(P\) will usually be skewed and
has a long tail.

@@

We tried to train a computer vision model to approximate this distance
measure with a residual plot. Let \(D\) be the result of equation
\ref{eq:kl-0}, and our estimator \(\hat{D}\) is formulated as

@@

With the approximated distance \(\hat{D}\), we will be able to know how
different the underlying distribution of the residuals is from a good
residual distribution. The approximated distance \(\hat{D}\) can also be
used as an index of the model violations. It also provides information
for the strength of the visual signal embedded in the residual plot.

@@

\hypertarget{approximated-distance-as-an-model-violations-index-mvi}{%
\subsubsection{Approximated distance as an Model Violations Index
(MVI)}\label{approximated-distance-as-an-model-violations-index-mvi}}

In Section \ref{output-formats}, we discuss the output formats of the
computer vision model presented in this paper. One notable advantage
highlighted is its interpretability. The approximated distance
\(\hat{D}\) is an estimator of the difference between the actual
residual distribution and the reference residual distribution. This
difference primarily arises from deviations in model assumptions. The
magnitude of \(D\) directly reflects the degree of these deviations,
thus making \(\hat{D}\) instrumental in forming a model violation index.

Note that since the Kullback-Leibler divergence could be affected by the
number of random variables involved in the evaluation. In general, more
observations will result in greater distance \(D\). This does not mean
\(\hat{D}\) can not correctly represent the extent of the model
violations since more observations in a residual plot usually lead to
stronger visual signal strength as the underlying visual patterns will
have higher chance to be revealed except for cases like serious
overlapping. For the same data generating process, \(D\) usually grows
logarithmically as the number of observations increases. This is because
\(D = \text{log}(1 + D_{KL})\), and
\(D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}\) under the independence
assumption.

Note that the Kullback-Leibler divergence might be influenced by the
number of random variables involved in its evaluation. Generally, a
larger number of observations will lead to a greater distance \(D\).
However, this does not imply that \(\hat{D}\) fails to accurately
represent the extent of model violations. In fact, when examining
residual plots with more observations, we often observe a stronger
visual signal strength, as the underlying patterns are more likely to be
revealed, except in cases of significant overlapping.

For a consistent data generating process, \(D\) typically increases
logarithmically with the number of observations. This behaviour comes
from the relationship \(D = \text{log}(1 + D_{KL})\), where
\(D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}\) under the assumption of
independence.

Therefore, the Model Violations Index (MVI) can be proposed as

\begin{equation}
\label{eq:mvi}
\text{MVI} = C + \hat{D} - log(n),
\end{equation}

\noindent where \(C\) is a large enough constant keeping the result
positive.

@@

Non-normality violations arise from specifying a non-normal distribution
for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
distinct error distributions are considered, including discrete,
uniform, normal, and log-normal distributions, as presented in Figure
\ref{fig:different-e}. Each distribution imparts unique characteristics
to the residuals. The discrete error distribution introduces
discreteness in residuals, while the log-normal distribution typically
yields outliers. Uniform error distribution may result in residuals
filling the entire space of the residual plot. All of these
distributions exhibit visual distinctions from the normal error
distribution.

@@

For each generated residual plot, we computed four scagnostics --
``Monotonic,'' ``Sparse,'' ``Splines,'' and ``Striped'' -- using the
\texttt{cassowaryr} R package \citep{mason2022cassowaryr}. These
computed measures, along with the number of observations from the fitted
model, were provided as the second input for the computer vision model.
Although other scagnostics are informative, they are currently
unavailable due to a fatal bug in the compiled C program of the
\texttt{interp} R package \citep{Albrecht2023interp}, which may
unpredictably crash the process. For reproducibility, we excluded these
scagnostics from the training data.

@@

The model begins with an input layer of shape
\(n \times h \times w \times 3\), capable of handling \(n\) RGB images.
This is followed by a grayscale conversion layer utilizing the luma
formula under the Rec. 601 standard, which converts the colour image to
grayscale. Grayscale suffices for our task since data points are plotted
in black. We experiment with three combinations of \(h\) and \(w\):
\(32 \times 32\), \(64 \times 64\), and \(128 \times 128\), aiming to
achieve sufficiently high image resolution for the problem at hand.

@@

To achieve an optimal deep learning model, hyperparameters like the
learning rate often need to be fine-tuned using a tuner. In our study,
we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
Python library \citep{omalley2019kerastuner} for this purpose. A
comprehensive list of hyperparameters is provided in Table
\ref{tab:hyperparameter}.

@@

This section remains incomplete as we may modify the ``Results'' section
later on.

@@

The figure highlights that most under-predictions occurred when
\(\hat{D} < 3\), while over-predictions occurred predominantly when
\(3 < \hat{D} < 6\). To comprehend the reasons behind these deviations,
we fitted a regression tree using the \texttt{rpart} R package
\citep{Terry2022rpart} on \(D - \hat{D}\) for all factors used in the
data generating process. The results of the regression tree are provided
in the Appendix.

The regression tree revealed that most issues arose from non-linearity
problems and the presence of a second predictor in the regression model.
When the error distribution had a very small variance, all computer
vision models tended to under-predict the distance. Conversely, when the
error distribution had a large variance, all models tended to
over-predict the distance. Therefore, Additionally, for input images
representing null plots, it was expected that the models will
over-predict the distance, as explained in Section
\ref{null-distribution-of-the-approximated-distance}.

@@

\hypertarget{model-violations-index-mvi}{%
\subsection{Model Violations Index
(MVI)}\label{model-violations-index-mvi}}

Figures \ref{fig:poly-index} and \ref{fig:heter-index} display the
residual plots for fitted models exhibiting varying degrees of
non-linearity and heteroskedasticity. Each residual plot's Model
Violation Index (MVI) is computed using Equation \ref{eq:mvi} with
\(C = 10\). When \(\text{MVI} > 8\), the visual patterns are notably
strong and easily discernible by humans. In the range
\(6 < \text{MVI} < 8\), the visibility of the visual pattern diminishes
as MVI decreases. Conversely, when \(\text{MVI} < 6\), the visual
pattern tends to become relatively faint and challenging to observe.
Table \ref{tab:mvi} provides a summary of the MVI usage and it is
applicable to other linear regression models.

\caption{\label{tab:mvi}Degree of model violations or the strength of the visual signals according to the Model Violation Index (MVI). The constant $C$ is set to be 10.}
\centering
\begin{tabular}[t]{ll}
\toprule
Degree of model violations & Range ($C$ = 10)\\
\midrule
Strong & $\text{MVI} > 8$\\
Moderate & $6 < \text{MVI} < 8$\\
Weak & $\text{MVI} < 6$\\
\bottomrule
\end{tabular}
\end{table}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/poly-index-1}

\caption{Residual plots generated from fitted models exhibiting varying degrees of non-linearity violations. The Model violations index (MVI) is displayed atop each residual plot. The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.}\label{fig:poly-index}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/heter-index-1}

\caption{Residual plots generated from fitted models exhibiting varying degrees of heteroskedasticity violations. The Model violations index (MVI) is displayed atop each residual plot. The heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.}\label{fig:heter-index}
\end{figure}

\hypertarget{comparison-with-human-visual-inference-and-conventional-tests}{%
\subsection{Comparison with human visual inference and conventional
tests}\label{comparison-with-human-visual-inference-and-conventional-tests}}

@@

For each lineup used in \citet{li2023plot}, there is one actual residual
plot and 19 null plots. While the distance \(D\) for the actual residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to predict \(\hat{D}\) for both the actual residual plots
and the null plots. Additionally, the appropriate conventional tests
including the Ramsey Regression Equation Specification Error Test
(RESET) \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
test \citep{breusch1979simple} for heteroskedasticity were applied on
the same data for comparison.

@@

Likewise, Figure \ref{fig:conv-mosaic} and Table \ref{tab:conv-table}
provide a summary of the agreement between decisions made by the
computer vision model and conventional tests. The agreement rates
between conventional tests and the computer vision model are 65.37\% and
57.29\% for residual plots containing heteroskedasticity and
non-linearity patterns, respectively. These figures are lower than those
presented in Table \ref{tab:human-table}, indicating that the computer
vision model exhibits behaviour more akin to visual tests than
conventional tests. Additionally, conventional tests tend to reject
residual plots almost exclusively when the computer vision model does,
but they also reject plots when the computer vision model does not. This
suggests that conventional tests are more sensitive than the computer
vision model.

When plotting the decision against the distance, as illustrated in
Figure \ref{fig:power}, several observations emerge. Firstly, for the
computer vision model, when the distance \(D > 3\), almost all residual
plots are rejected. However, for visual tests conducted by humans, the
threshold is higher, set at \(D > 5\). And conventional tests exhibit
some non-rejection cases for \(D > 3\). Moreover, the computer vision
model tends to have fewer rejected cases than conventional tests when
\(D < 2\). However, visual tests demonstrate the lowest sensitivity to
residual plots with small distances. Therefore, tests based on the
computer vision model are less sensitive to small deviations from model
assumptions than conventional tests but more sensitive to moderate
deviations. Rejection decisions are fitted by logistic regression models
with no intercept terms but an offset equals to
\(\text{log}(0.05/0/95)\). The fitting curves for the computer vision
model fall between those of conventional tests and visual tests for
non-linearity and are slightly below conventional tests for
heteroskedasticity. There is still potential to refine the computer
vision model to better align its behaviour with visual tests.

@@

Figure \ref{fig:delta} indicates that the weighted detection rate
increases as the \(\delta\)-difference increases, particularly when the
\(\delta\)-difference is greater than zero. A negative
\(\delta\)-difference suggests that there is at least one null plot in
the lineup with a stronger visual signal than the actual residual plot.
In some instances, the weighted detection rate is close to one, yet the
\(\delta\)-difference is negative. This discrepancy implies that the
distance measure, or the approximated distance, may not perfectly
reflect actual human behaviour.

@@

\caption{\label{tab:human-table}Summary of decisions made by visual tests conducted by human and decisions made by the computer vision model.}
\centering
\begin{tabular}[t]{llll}
\toprule
Violations & \#Samples & \#Agreements & Agreement rate\\
\midrule
heteroskedasticity & 540 & 367 & 0.6796\\
non-linearity & 576 & 385 & 0.6684\\
\bottomrule
\end{tabular}
\end{table}

\caption{\label{tab:conv-table}Summary of decisions made by conventional tests and decisions made by the computer vision model.}
\centering
\begin{tabular}[t]{llll}
\toprule
Violations & \#Samples & \#Agreements & Agreement rate\\
\midrule
heteroskedasticity & 540 & 464 & 0.6537\\
non-linearity & 576 & 459 & 0.5729\\
\bottomrule
\end{tabular}
\end{table}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/conv-mosaic-1}

\caption{Rejection rate ($p$-value $\leq0.05$) of conventional test conditional on the computer vision model decision on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. The conventional test rejects more frequently than the computer vision model, and (almost) only rejects when the computer vision model does.}\label{fig:conv-mosaic}
\end{figure}

@@

\caption{Comparison of decisions made by visual tests, conventional tests and the computer vision model. The data points are fitted with logistic regression models with no intercept but an offset equals to $\text{log}(0.05/0.95)$.}\label{fig:power}
\end{figure}

@@

In this section, we will utilize the trained computer vision model on
three example datasets. These include the dataset associated with the
residual plot displaying a ``left-triangle'' shape, as displayed in
Figure \ref{fig:false-finding}, along with the Boston housing dataset,
and the ``dino'' datasets from the \texttt{datasaurus} R package.
Additionally, we will outline the workflow for employing this model with
the \texttt{autovi} R package.

\hypertarget{left-triangle}{%
\subsubsection{Left-triangle}\label{left-triangle}}

In Section \ref{introduction}, we presented an example residual plot
showcased in Figure \ref{fig:false-finding}, illustrating how humans
might misinterpret the ``left-triangle'' shape as indicative of
heteroskedasticity. Additionally, the Breusch-Pagan test yielded a
rejection with a \(p\)-value of 0.046, despite the residuals originating
from a correctly specified model. Figure \ref{fig:false-lineup} offers a
lineup for this fitted model, showcasing various degrees of
``left-triangle'' shape across all residual plots. This phenomenon is
evidently caused by the skewed distribution of the fitted values.
Notably, if the residual plot were evaluated through a visual test, it
would not be rejected since the actual residual plot positioned at 10
can not be distinguished from the others.

Moving forward, we will employ the trained computer vision model to
assess the residual plot with the assistance of the \texttt{autovi} R
package. This package is accessible on the GitHub repository
\texttt{TengMCing/autovi}, and users can install it by executing the
command \texttt{remotes::install\_github("TengMCing/autovi")} in R.

The results of the \texttt{list\_keras\_model()} function provide a list
of available trained computer vision models. To perform the residual
plot assessment, users need to supply a fitted linear regression model
and select a trained Keras model to initialize the checker.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checker }\OtherTok{\textless{}{-}} \FunctionTok{auto\_vi}\NormalTok{(}\AttributeTok{fitted\_mod =}\NormalTok{ mod, }
                   \AttributeTok{keras\_mod =} \FunctionTok{get\_keras\_model}\NormalTok{(}\StringTok{"vss\_phn\_32"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

When invoking the \texttt{check()} method to execute the assessment,
users can specify the number of null plots and bootstrapped samples
using the \texttt{null\_draws} and \texttt{boot\_draws} arguments,
respectively. The diagnostic results can be viewed by directly printing
the object. Alternatively, a summary plot of the assessment can be
generated using the \texttt{summary\_plot()} method.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checker}\SpecialCharTok{$}\FunctionTok{check}\NormalTok{(}\AttributeTok{null\_draws =}\NormalTok{ 200L, }\AttributeTok{boot\_draws =}\NormalTok{ 200L)}
\NormalTok{checker}\SpecialCharTok{$}\FunctionTok{summary\_plot}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

Figure \ref{fig:false-check} presents the results of the assessment.
Notably, the observed visual signal strength is considerably lower than
the 95\% sample quantile of the null distribution. Moreover, the
bootstrapped distribution suggests that it is highly improbable for the
fitted model to be misspecified as the majority of bootstrapped fitted
models will not be rejected.

Thus, for this particular fitted model, both the visual test and the
computer vision model will not reject \(H_0\). However, the
Breusch-Pagan test will reject \(H_0\) because it can not effectively
utilize information from null plots.

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-check-1}

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. On the left, the residual plot exhibiting "left-triangle" shape is displayed. The blue area indicates the distribution of approximated distances for null plots, while the red area represents the distribution of approximated distances for bootstrapped plots. The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$.}\label{fig:false-check}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-lineup-1}

\caption{A lineup of residual plots displaying "left-triangle" visual patterns. The actual residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.}\label{fig:false-lineup}
\end{figure}

\hypertarget{boston-housing}{%
\subsubsection{Boston housing}\label{boston-housing}}

The Boston housing dataset, originally published by
\citet{harrison1978hedonic}, offers insights into housing in the Boston,
Massachusetts area. For illustration purposes, we will utilize a reduced
version from Kaggle, comprising 489 rows and 4 columns: average number
of rooms per dwelling (RM), percentage of lower status of the population
(LSTAT), pupil-teacher ratio by town (PTRATIO), and Median value of
owner-occupied homes in \$1000's (MEDV). In our analysis, MEDV will
serve as the response variable, while the other columns will function as
predictors in a linear regression model. Our primary focus is to detect
non-linearity, because the relationships between RM and MEDV or LSTAT
and MEDV are non-linear.

Figure \ref{fig:boston-check} displays the residual plot and the
assessment conducted by the computer vision model. A clear non-linearity
pattern resembling a ``U'' shape is shown in the residual plot.
Furthermore, the RESET test yields a very small \(p\)-value. The
approximated distance \(\hat{D}\) significantly exceeds
\(Q_{null}(0.95)\), leading to rejection of \(H_0\). The bootstrapped
distribution also suggests that almost all the bootstrapped fitted
models will be rejected, indicating that the fitted model is unlikely to
be correctly specified.

Additionally, the lineup of this fitted model presented in Figure
\ref{fig:boston-lineup} underscores that the ``U'' shape is a
distinctive characteristic of the actual residual plot, absent in any of
the null plots. Consequently, if a visual test is conducted, \(H_0\)
will also be rejected.

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-check-1}

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. On the left, the residual plot of the Boston housing fitted model is displayed. The blue area indicates the distribution of approximated distances for null plots, while the red area represents the distribution of approximated distances for bootstrapped plots. The fitted model will be rejected since $\hat{D} >= Q_{null}(0.95)$.}\label{fig:boston-check}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-lineup-1}

\caption{A lineup of residual plots for the Boston housing fitted model. The actual residual plot is at position 7. It can be easily identified as the most different plot.}\label{fig:boston-lineup}
\end{figure}

@@

The computer vision model possesses the capability to detect not only
typical issues like non-linearity, heteroskedasticity, and non-normality
but also artifact visual patterns resembling real-world objects, as long
as they do not appear in null plots. These visual patterns can be
challenging to categorize in terms of model violations. Therefore, we
will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk
test \citep{shapiro1965analysis} for comparison.

The ``dino'' dataset within the \texttt{datasaurus} R package
exemplifies this scenario. With only two columns, x and y, fitting a
regression model to this data yields a residual plot resembling a
``dinosaur,'' as displayed in Figure \ref{fig:dino-check}.
Unsurprisingly, this distinct residual plot stands out in a lineup, as
shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans
would undoubtedly reject \(H_0\).

According to the residual plot assessment by the computer vision model,
\(\hat{D}\) exceeds \(Q_{null}(0.95)\), warranting a rejection of
\(H_0\). Additionally, most of the bootstrapped fitted models will be
rejected, indicating an misspecified model. However, both the RESET test
and the Breusch-Pagan test yield \(p\)-values greater than 0.3, leading
to a non-rejection of \(H_0\). Only the Shapiro-Wilk test rejects the
normality assumption with a small \(p\)-value.
